---
title: "Single-cell deconvolution methods EMBL-EBI course"
author: "Estef Vazquez & Gabriel Pereira"
output:
  html_document:
    theme: simplex
    toc: yes
    toc_float:
      collapsed: true
---

```{r, setup, include=FALSE}

# Load Required Libraries
library(here)
library(tidyverse)
library(plotly)
library(ggthemes)
library(ggpubr)
library(RColorBrewer)
library(colorspace)
library(cowplot)
library(Biobase)
library(BisqueRNA)
suppressWarnings(library(BayesPrism))
suppressWarnings(library(Seurat))

# Double check your working directory
getwd()

# Set global chunk options
knitr::opts_chunk$set(
  echo = TRUE,              
  message = FALSE,         
  warning = FALSE,         
  error = FALSE,            
  fig.width = 10,           
  fig.height = 6,          
  out.width = "100%",       
  cache = TRUE,
  dev = 'png',
  fig.path = here("figures", "/")
)

# Create directories 
dir.create(here("data"), recursive = TRUE, showWarnings = FALSE)
dir.create(here("results"), recursive = TRUE, showWarnings = FALSE)
dir.create(here("figures"), recursive = TRUE, showWarnings = FALSE)
message("Project directories created at:")
message("Data directory: ", here("data"))
message("Results directory: ", here("results"))
message("Figures directory: ", here("figures"))

```

This vignette provides a practical overview of second generation methods for cell-type deconvolution using transcriptomic data. We'll explore some commonly used approaches:

1.  **Bisque**: A reference-based method that uses single-cell data to deconvolve bulk samples
2.  **BayesPrism**: A Bayesian framework that accounts for cell state transitions
3.  **CIBERSORTx**: A support vector regression approach
4.  **InferCNV**: A method to identify copy number variations from transcriptomic data

For these exercises, we'll be utilizing Pancreatic Ductal Adenocarcinoma (PDAC) data from the TCGA project, which has been downsampled for computational efficiency.


## Data Loading

Let's start by loading and examining our datasets: 1) bulk RNA-seq data, 2) single-cell reference data, and 3) clinical information.

```{r load_data, cache=FALSE}

# Load RDS files 
read_rds <- function(file_path, default = NULL) {
  tryCatch({
    message(paste("Loading", file_path))
    readRDS(file_path)
  }, error = function(e) {
    message(paste("Error loading", file_path, ":", e$message))
    return(default)
  })
}

# Load bulk data - raw counts 
bulkdata <- read_rds("data/tcga_pdac_counts.rds")
if (!is.null(bulkdata)) {
  bulkdata <- as.matrix(bulkdata)
  message(paste("Bulk data dimensions:", paste(dim(bulkdata), collapse=" x ")))
}


# Load TPMs
tpms <- read_rds("data/tcga_pdac_tpm.rds")
if (!is.null(tpms)) {
  tpms <- tpms %>% rownames_to_column(var = "Gene")
  message(paste("TPM data dimensions:", paste(dim(tpms), collapse=" x ")))
}


# Load single-cell Seurat object
sobj <- read_rds("data/sc_ref_pdac_seurat.rds")
if (!is.null(sobj)) {
  message("Single-cell reference data:")
  print(sobj)
  message("Available layers:")
  print(Layers(sobj))
}


# Load clinical data
clindata <- read_rds("data/tcga_pdac_clindata.rds")
if (!is.null(clindata)) {
  message("Clinical data columns:")
  print(names(clindata))
}

```

## Exercise 1. Understanding Input Data

-Identify each data object and determine its dimensions

-How many individual cells are in the single-cell reference?

-Which clinical variables are available? Is the clinical data complete?

-Take a moment to explore each object


## Exercise 2. Bisque

Bisque (Bulk RNA-seq Imputation from Single-cell Expression) is a reference-based deconvolution method that leverages single-cell RNA-seq data to estimate cell-type proportions in bulk tissue samples. It accounts for variations within cell types and cross-subject heterogeneity.

Preparing Input

```{r bisque_preparation}

# Prepare the bulk data as an ExpressionSet object
# Bisque expects features as rows and samples as columns
bulk.eset <- ExpressionSet(assayData = bulkdata)


# Extract raw counts from the Seurat object
sc_counts <- GetAssayData(sobj, layer = "counts")


#  Prepare single-cell metadata 
cellmeta <- sobj@meta.data
celltype <- cellmeta[, c("cell_type", "patient")]
colnames(celltype) <- c("cellType", "SubjectName") #renaming


# Create AnnotatedDataFrame object for phenotype data
df <- data.frame(labelDescription=colnames(celltype), row.names = colnames(celltype))
sc_annot <- new("AnnotatedDataFrame", data=celltype, varMetadata=df)


# Clean up 
rm(cellmeta, celltype, df)
gc()

# Create ExpressionSet object for single-cell reference
sc.eset <- ExpressionSet(assayData = as.matrix(sc_counts), phenoData = sc_annot)


# Clean up large objects to free memory
rm(sc_counts, sc_annot, sobj)
gc()


# Running Bisque Deconvolution 
res <- BisqueRNA::ReferenceBasedDecomposition(bulk.eset, 
                                              sc.eset, 
                                              markers=NULL, 
                                              use.overlap=FALSE)
gc()
res

# Explore results
bisque_results <- as.data.frame(res$bulk.props)
bisque_results <- t(bisque_results) #transpose for visualization


# Save
saveRDS(bisque_results, 
        here("results", "bisque_results.rds"))


```

Questions

1.  How many cells are in the single-cell reference?

2.  What were the most and least abundant cell types present in the data results? visualize Bisque results.

3.  How might the heterogeneity within PDAC samples affect the deconvolution results? What biological features of PDAC make it challenging for cell type deconvolution?


Hint to question 2

```{r visualize_bisque, fig.height=8}

# Format data 
bisque_long <- bisque_results %>%
  as.data.frame() %>%
  rownames_to_column("SampleID") %>%
  pivot_longer(-SampleID, names_to = "CellType", values_to = "Proportion") %>% glimpse()


# Boxplot of cell type proportions
ggplot(bisque_long, aes(x=reorder(CellType, Proportion, FUN=median), y=Proportion, fill=CellType)) +
  geom_boxplot(outlier.shape = 21, outlier.alpha = 0.5) +
  geom_jitter(width=0.2, size=1, alpha=0.5) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle=45, hjust=1),
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold")
  ) +
  labs(
    title = "Cell Type Proportions Estimated by Bisque",
    x = "Cell Type",
    y = "Estimated Proportion"
  ) +
  scale_fill_brewer(palette = "Set3")


# Summary statistics
bisque_summary <- bisque_long %>%
  group_by(CellType) %>%
  summarize(
    Mean = mean(Proportion),
    Median = median(Proportion),
    SD = sd(Proportion),
    Min = min(Proportion),
    Max = max(Proportion)
  ) %>%
  arrange(desc(Mean)) %>% 
  glimpse()


# Display table
knitr::kable(bisque_summary, caption = "Summary of Cell Type Proportions", digits = 3)

```

## Exercise 3: Deconvolution using BayesPrism

```{r}
# BayesPrism expects samples as rows and genes as columns (opposite of Bisque) - (sample x gene)
bulkdata <- as.data.frame(t(bulkdata))

# Explore object
dim(bulkdata)
head(rownames(bulkdata)) #samples as rownames
head(colnames(bulkdata)) #genes as colnames

# clinical data
glimpse(clindata)

# Seurat object
sobj
#colnames(sobj)
#nrow(sobj) #18.5k
#Features(sobj)
#Layers(sobj)

# Retrievesingle-cell metadata
cellmeta <- sobj@meta.data

# Rename cell_meta columns
cell_meta <- cellmeta %>% dplyr::rename("cell.type.labels" = "cell_type",
                                         "cell.state.labels" = "cell_state") 
glimpse(cell_meta)


# If you have different levels of granularity for cell annotation, you can use them But, rare cell types or cell states might be challenging to estimate

# As a start we will run at a higher level of cell annotation
cell.type.labels <- cell_meta$cell.type.labels
cell.state.labels <- cell_meta$cell.state.labels
cell.type.labels <- as.character(cell.type.labels)
cell.state.labels <- as.character(cell.state.labels)


# cell.type.labels will be a character vector of the same length as nrow(sc.dat)
# Explore the cell type labels
sort(table(cell.type.labels))

# cell.state.labels is a character vector of the same length as nrow(sc.dat) 
sort(table(cell.state.labels))

# Create df with cell annotation
cell_annot <- as.data.frame(table(cbind.data.frame(cell.state.labels, cell.type.labels)))
glimpse(cell_annot)


# Extract the counts from the seurat object
sc.dat <- GetAssayData(sobj, assay='RNA', slot='counts')
dim(sc.dat)



# Convert the sparse matrix to a dense matrix
sc.dat <- as.matrix(sc.dat) 
gc()
colnames(sc.dat)
rownames(sc.dat)


# Before deconvolution, you can check for correlation between cell types
# Plot pairwise correlation matrix 
plot.cor.phi (input=sc.dat,
              input.labels=cell.type.labels,
              title="cell state correlation",
              pdf.prefix = "cor_matrix",
              cexRow=0.5, cexCol=0.5,
              margins=c(4,4)
)


plot.cor.phi (input=sc.dat,
              input.labels=cell.state.labels,
              title="cell state correlation",
              pdf.prefix = "cor_matrix_cell_states",
              cexRow=0.4, cexCol=0.4,
              margins=c(5,5)
)



# Visualize distribution of outlier genes from scRNA-seq reference
sc.stat <- plot.scRNA.outlier(
  input=sc.dat, 
  cell.type.labels=cell.type.labels,
  species="hs", 
  return.raw=TRUE, 
  pdf.prefix = "outlier_sc"
)


# Visualize outlier genes from bulk RNA-seq
bk.stat <- plot.bulk.outlier(
  bulk.input=bulkdata, 
  sc.input=sc.dat, #make sure colnames are gene symbol or ENSEMBL ID
  cell.type.labels=cell.type.labels,
  species="hs", 
  return.raw=TRUE,
  pdf.prefix = "outlier_bulk" 
)



# Filter matrix removing features that add noise 
sc.dat.filtered <- cleanup.genes(input=sc.dat,
                                 input.type="count.matrix",
                                 species="hs",
                                 gene.group=c( "Rb","Mrp","other_Rb","chrM","MALAT1","chrX","chrY") ,
                                 exp.cells=5)

dim(sc.dat)
dim(sc.dat.filtered)


#free memory 
gc()


# Plot correlation between the features present in single-cell reference and bulk samples
plot.bulk.vs.sc(sc.input = sc.dat.filtered,
                bulk.input = bulkdata, 
                pdf.prefix = "concordance_sc_bulk")
gc()


# Subset the reference to retain protein coding genes only
sc.dat.filtered.pc <-  select.gene.type(sc.dat.filtered, gene.type = "protein_coding")
dim(sc.dat.filtered.pc)


# Create PRISM object
myPrism <- new.prism(
  reference=sc.dat.filtered.pc,
  mixture=bulkdata,
  input.type="count.matrix",
  cell.type.labels = cell.type.labels,
  cell.state.labels = cell.type.labels,
  key='Malignant',
  outlier.cut=0.01,
  outlier.fraction=0.1,
)
gc()


# Run BayesPrism deconvolution 
#?run.prism 
bp.res <- run.prism(prism = myPrism, n.cores=10) # takes approximately 6 minutes
bp.res


# Retrieve cell type fractions
bayes_results <- get.fraction(bp.res, which.theta = 'final', state.or.type = 'type')
bayes_results_df <- as.data.frame(bayes_results)

# Save
saveRDS(bayes_results_df, 
        here("results", "bayesprism_results.rds"))

write_csv(bayes_results_df, 
        here("results", "bayesprism_results.csv"))


# Transpose if required
bayes_res <- t(bayes_results_df)
bayes_res <- as.data.frame(bayes_res)


```

Questions

1.  Did you get an error? Why is the transposition step important?

2.  Which are the main cell types that are represented in the single-cell reference?

3.  You visualized correlations between cell types. How would you interpret high correlations between certain cell types, and how might this affect deconvolution accuracy?

4.  BayesPrism specifically asks for a "key" malignant cell type. How does this approach differ from other deconvolution methods, and what advantages might it offer when studying tumor samples?


## Exercise 4: CIBERSORTx

Log in to your account at <https://cibersortx.stanford.edu/> and select option 2: Impute Cell Fractions. Then, follow the instructions to input TPMs. Generally, you will need to do this steps:

1.  Formatting your input file

2.  Uploading your files (pulldown menu): the signature matrix (LM22 for this example) and your bulk expression file ("tpms_cibersortx.tsv"). Press the green button “+ add files”, find the files, name them and select the file type.

3.  Running CIBERSORTx: menu –\> Run CIBERSORTx. Click option 2. "Impute Cell Fractions”. For the Mixture file select your tpms_cibersortx.tsv file (you can find it in the github repository).  Run and download your output file. You can make statistics and further analyse these results.



## Exercise 5: CNV and cell type composition relationship

```{r}
# We will use pre-computed CNV scores (inferCNV)
# Load scores
cnvs <- read_csv("scref_metadata_new.csv")
glimpse(cnvs)


# Summary statistics of CNV scores by cell type
cnv_summary <- cnvs %>%
  group_by(cell_type) %>%
  summarize(
    mean_cnv = mean(cnv_score),
    median_cnv = median(cnv_score),
    sd_cnv = sd(cnv_score),
    n = n()
  ) %>%
  arrange(desc(mean_cnv))

print(cnv_summary)


# UMAP plot
ggplot(cnvs, aes(x = UMAP1, y = UMAP2, color = cnv_score)) +
  geom_point(size = 0.5, alpha = 0.7) +
  scale_color_viridis_c() +
  theme_minimal() +
  labs(title = "CNV Score Distribution on UMAP", color = "CNV Score")


# Subset malignant cells  
malignant_cells <- cnvs %>% 
  filter(grepl("Malignant", New_label))


#  Visualize CNV within malignant cells
ggplot(malignant_cells, aes(x = UMAP1, y = UMAP2, color = New_label)) +
  geom_point(size = 1) +
  theme_classic() + 
  labs(title = "High vs Low CNV Malignant Cells on UMAP", color = "CNV Group")


# Now we will link CNV patterns with cellular composition
glimpse(cnvs)
glimpse(bayes_results_df)
bayes_results_df <- bayes_results_df %>% rownames_to_column(var = "sample") %>% glimpse()


# Calculate average CNV scores by cell type 
cell_type_cnv <- cnvs %>%
  group_by(cell_type) %>%
  summarize(
    mean_cnv = mean(cnv_score),
    median_cnv = median(cnv_score),
    sd_cnv = sd(cnv_score),
    n_cells = n()
  )


# Using deconvolution results, calculate average proportions of each cell type across samples
bulk_cell_prop <- bayes_results_df %>%
  select(-sample) %>%
  summarize(across(everything(), list(mean = mean, sd = sd))) %>%
  pivot_longer(
    everything(),
    names_to = c("cell_type", ".value"),
    names_pattern = "(.+)_(.+)"
  )

# Take a look
glimpse(cell_type_cnv)
glimpse(bulk_cell_prop)


# Merge datasets
integrated_cell_types <- cell_type_cnv %>%
  left_join(bulk_cell_prop, by = "cell_type") %>% 
  glimpse()


# Visualize CNV vs cell type abundance relationship
ggplot(integrated_cell_types, aes(x = mean_cnv, y = mean)) +
  geom_point(aes(size = n_cells), alpha = 0.7) +
  geom_text(aes(label = cell_type), vjust = -1, size = 3) +
  geom_smooth(method = "lm", se = TRUE, color = "blue", alpha = 0.2) +
  scale_size_continuous(name = "Number of cells in reference") +
  theme_classic() +
  labs(title = "Relationship Between Cell Type CNV and Cell Type Abundance",
       x = "Mean CNV Score in Single-cell Data",
       y = "Mean Cell Type Proportion in Bulk Data") +
  # Add correlation statistics
  annotate("text", x = min(integrated_cell_types$mean_cnv), 
           y = max(integrated_cell_types$mean),
           label = paste("Pearson correlation:", 
                         round(cor(integrated_cell_types$mean_cnv, 
                                   integrated_cell_types$mean, 
                                   method = "pearson"), 3)),
           hjust = 0, vjust = 1)
```


Questions:

1. Do you observe a relationship between CNV burden and cellular composition? Does that make biological sense?

2. Why stromal cells might be abundant despite lower CNV scores?



## Optional: Explore the Omnideconv package

https://github.com/omnideconv/omnideconv
```{r}
library(omnideconv)
deconvolution_methods
```


## Github repository

Find more resources (references, links and vignettes) at the [**Github repository**](https://github.com/estefvazquez/sc_Deconvolution_course){.uri}**.**

